---
title: "Introduction to droughtwatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to droughtwatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
library(droughtwatch)
```

## Overview

**droughtwatch** is an R package for agricultural drought monitoring and early
warning. It provides:

- Standard drought indices (SPI, SPEI, VHI)
- Drought episode detection and frequency analysis
- NDVI-based vegetation monitoring
- Automated alerts and monthly bulletins
- An interactive Shiny dashboard

---

## 1. Load Example Data

```{r load-data}
data(morocco_climate)
head(morocco_climate)
```

The `morocco_climate` dataset contains 24 years of simulated monthly climate
data (precipitation, temperature, ETo) for a semi-arid Moroccan region.

---

## 2. Import Climate Data

```{r import}
climate <- import_climate_data(data = morocco_climate)
str(climate)
```

`import_climate_data()` handles CSV and Excel files, date parsing, and
missing value imputation. Here we pass a data frame directly.

---

## 3. Compute SPI

The **Standardized Precipitation Index (SPI)** quantifies precipitation
anomalies at different time scales.

```{r spi}
spi_result <- calc_spi(climate, scales = c(1, 3, 6))
head(spi_result[, c("date", "precip", "spi_3", "class_3")])
```

```{r spi-plot}
plot_drought_timeseries(spi_result, index_col = "spi_3", interactive = FALSE)
```

The orange, red, and dark-red dashed lines mark moderate (−1.0), severe (−1.5),
and extreme (−2.0) drought thresholds.

---

## 4. Compute SPEI

The **SPEI** adds evapotranspiration to the water balance, making it more
sensitive to temperature-driven droughts.

```{r spei}
spei_result <- calc_spei(climate, scales = c(1, 3), lat = 32)
head(spei_result[, c("date", "spei_3", "class_3")])
```

---

## 5. Detect Drought Episodes

```{r detect}
drought_res <- detect_drought(spi_result, index_col = "spi_3", threshold = -1.0)
drought_res$episodes
```

Each row describes one drought episode: start/end dates, duration in months,
mean intensity, and alert level.

---

## 6. Frequency Analysis

```{r frequency}
stats <- analyze_frequency(drought_res, by = "year")
stats$summary
stats$severity_table
```

---

## 7. Generate Alerts

```{r alerts}
alerts <- generate_alerts(drought_res, region = "Beni Mellal-Khenifra")
# Show only active alerts
alerts[alerts$alert_level != "NONE", ]
```

---

## 8. Forecast

```{r forecast}
fc <- forecast_drought(spi_result, index_col = "spi_3",
                        method = "moving_average", horizon = 3)
fc
```

---

## 9. NDVI & VHI (spatial analysis)

```{r ndvi, eval = requireNamespace("terra", quietly = TRUE)}
# Generate a synthetic NDVI stack
ndvi <- download_ndvi("2020-01-01", "2020-06-30",
                       region = c(-6, 31, -4, 33),
                       source = "synthetic")

# Compute VHI from NDVI
vhi <- calc_vhi(ndvi)

# Map June 2020
plot_drought_map(vhi[[6]], type = "vhi", title = "VHI - June 2020")
```

---

## 10. Generate Monthly Bulletin

```{r bulletin, eval = FALSE}
generate_monthly_bulletin(
  index_data    = spi_result,
  drought_result = drought_res,
  alerts        = alerts,
  region        = "Beni Mellal-Khenifra",
  output_format = "html"
)
```

---

## 11. Interactive Dashboard

```{r shiny, eval = FALSE}
shiny_dashboard(data = morocco_climate, region = "Beni Mellal-Khenifra")
```

---

## References

- McKee, T.B., Doesken, N.J. & Kleist, J. (1993). The relationship of drought
  frequency and duration to time scales. *Proceedings of the 8th Conference on
  Applied Climatology*, 17–22 January, Anaheim, California.
- Vicente-Serrano, S.M., Beguería, S. & López-Moreno, J.I. (2010). A
  multiscalar drought index sensitive to global warming: the SPEI. *Journal of
  Climate*, 23(7), 1696–1718.
- Kogan, F.N. (1995). Application of vegetation index and brightness temperature
  for drought detection. *Advances in Space Research*, 15(11), 91–100.
